### 2018-07-18 source/Ch7-3C295/3c295.tex
final commit for sake of backup - thesis completed

\pg
The resulting model is expected to be reliable enough to calibrate our Groth Strip data for the full LOFAR array and the entire HBA bandwidth. Direction-dependent gains can then be used for baselines with only core and remote stations, but not those with international stations - direction-dependent effects are expected to arise in the final images due to this. Quantifying the impact of this problem is the subject of \cref{section.decorr}.

@@

\pg
As outlined in \cref{section.RIME}, calibration is the process of solving for Jones matrices, which are associated to individual antennas. Here, the only source in our model (the source that dominates, by far, over other sources in the field) is 3C295; we can thus limit ourselves to \textit{direction-independent} calibration, which is equivalent to performing direction-dependent calibration in the direction of 3C295 only. The associated RIME can be written as:
\begin{align}
\Vis_{pq}^{\nu} &= \Gjones_p^{\nu} \left( \sum_{s=3C295} \Ejones_{sp}^{\nu} \Kjones_{sp}^{\nu} \Bmatrix \left(\Kjones_{sq}^{\nu}\right)^H \left(\Ejones_{sq}^{\nu}\right)^H \right) \left(\Gjones_q^{\nu}\right)^H\\
		  &= \Gjones_p^{\nu} \Ejones_{3C295,p}^{\nu} \Bmatrix_{3C295}^{\nu} \left(\Ejones_{3C295,q}^{\nu} \Gjones_q^{\nu}\right)^H\\
		  &= \Jones_{3C295,p}^{\nu} \Bmatrix_{3C295}^{\nu} \left(\Jones_{3C295,q}^{\nu}\right)^H\label{3c295.dirin.cal}
\end{align}
where we mute the time-dependence of all terms, and explicitly model the LOFAR beam during both calibration and imaging, so as to ensure that the true sensitivity of each baseline to this out-of-phase-centre calibrator is taken into account appropriately.

\pg
What interests us here is therefore the $\Jones_{3C295,p}^{\nu}$ for all international stations $p$ and at all times and frequencies $t,\nu$. We improve conditioning by using time and frequency solution intervals of a minute and 8 subbands; at those times where this interval is too large, the quality-based weighting scheme is expected to compensate. Choosing such large intervals also helps self-calibration converge faster. We do not explicitly resort to physics-based solutions to constrain the calibration solutions: we solve for the full 8 parameters of each complex $2\times 2$ Jones matrix.

@@

\pg
Because we apply a $uv$-cut of 1km when calibrating, and a $uv$-cut of 100km when imaging, we need to make sure that we do not introduce spurious structure in the core and remote station gains. Similarly, we can check the quality of our self-calibration by looking at the structure of the international station gains. Gain structure can consist of genuine atmospheric \& instrumental effects, but can also include unmodeled flux being absorbed into the gain solutions. At first, this is unavoidable: if flux is resolved by a baseline but not included in the model, then it will be absorbed into the gains. As we improve our model, we expect the gain curves to flatten and become smoother. If this happens for our international station gains, then we have a strong indication that we are in fact recovering physical signal rather than adding artefacts into our model. Similarly, we expect core and remote station gains to remain mostly static: if this is the case, then our strategy does not bias our gain solutions for these stations away from the underlying truth.

@@

In so doing, we introduce a frequency-dependent error factor in our gain solutions:
\begin{align}
\hatJones_{3C295,p}^{\nu} &\approx \sqrt{\alpha_\nu} \Jones_{3C295,p}^{\nu}
\end{align}
where $\hatJones_{3C295,p}^{\nu}$ is our estimation of the Jones matrix $\Jones_{3C295,p}^{\nu}$ for time $t$ and frequency $\nu$. $\alpha_\nu$ is a smoothly-varying function of frequency alone. This then introduces a frequency-dependent error factor $1/\sqrt{\alpha_\nu}$ in our visibilities and therefore in our images. This factor is a constant for all visibilities at a given frequency.

@@

% images:
% IMAGES/3c295.fromVLA.6sb.withbeam.pass[1-5].int.model.fits IMAGES/3c295.fromVLA.60sb.withbeam.pass[1-3].int.model.fits IMAGES/MSlist.fullBW.LOBOS6.MSlist.fullBW.txt.uniform.int.model.fits
% script: /home/offler/plotSCamp.py
\begin{figure}[h!]
\includegraphics[width=0.8\linewidth]{images/AmpFluct.png}
\caption{\label{fig.SC.ampfluct} Amplitude fluctuation of the models resulting from self-calibration iterations, expressed as a ratio of the model's integrated flux divided by the Scaife \& Healde integrated flux model at the appropriate frequency. Note that the integrated model flux for 324sb-pass1 (the penultimate data point in this plot) was lost to a careless command: the value shown here is just the average of the two neighbouring values.}
\end{figure}

\pg
Because $\alpha$ is smooth in frequency, we can apply a frequency-dependent amplitude correction (i.e. ``bootstrap") by relying on the work of Scaife \& Heald \citep[see][]{arse}. This correction will be done in the imaging plane, since $alpha$ is a function of $uv$-position (but not sky position). This is, again, because 3C295 does not lie at phase centre: if it did, the correction could be done equivalently in visibility or image space. By performing it in the image-plane, we rid ourselves of the $uv$-dependence. The resulting signal-to-noise improvement is shown in \cref{fig.SCrms}. It is calculated on the bootstrapped residual images resulting from each round of the self-calibration process - as such, the only thing that changes is the noise, since the signal is corrected to be the appropriate constant As we can see, the noise falls nicely with self-calibration passes. Tragically, the results from the first full-bandwidth calibration run with the 60-subband model were lost before this plot could be made, and so only the results from the second full-bandwidth self-calibration pass remain. We can see that it is a small improvement over the 60-subband model. 
\begin{figure}[h!]
\includegraphics[width=0.8\linewidth]{images/SCrms.png}
\caption{\label{fig.SCrms} Bootstrapped residual RMS for each calibration pass. Note that the data for 324sb-pass1 (the penultimate data in this plot) were lost to a careless command: the value shown here is just the average of the two neighbouring values.}
\end{figure}

\pg
The large decrease (factor of $\sim2$)in RMS between the last 6-subband residual image and the first 60-subband residual image in \cref{fig.SCrms} is significant: we would expect a similar jump when going from 60 to 324 subbands. The fact that we do not is a strong indicator that we are digging into the limits of what our strategy can achieve by the end of the second full-bandwidth self-calibration pass: we are stuck in a local minimum. This is likely due to the fact that our initial sky model (and all subsequent models) only include 3C295: the fact that all the other sources in the primary beam are not included in the model is bound to introduce bias our gain solutions and artefacts in our images.

@@

\pg
Let us begin by checking whether our core and remote station gains change over time due to our self-calibration procedure. Here, we look at the gains of SB155, on which we solved for gain solutions using our initial and final models. This subband was included in the 60-subband self-calibration step, but not the 6-subband self-calibration step.
\begin{figure}[h!]
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.CS005HBA0.firstpass}.png}
\caption{\label{fig.sb155.gains.cs201.before} CS005, first calibration}
\end{subfigure}
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.CS005HBA0.lastpass}.png}
\caption{\label{fig.sb155.gains.cs201.after} CS005, last calibration}
\end{subfigure}
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.RS305HBA.firstpass}.png}
\caption{\label{fig.sb155.gains.RS305.before} RS305, first calibration}
\end{subfigure}
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.RS305HBA.lastpass}.png}
\caption{\label{fig.sb155.gains.RS305.afer} RS305, last calibration}
\end{subfigure}
\caption{\label{fig.sb155.gains.CR} Gain amplitude \& phase for CS201 of SB155}
\end{figure}

@@

\pg
So far, we have shown the gain solutions for one subband, out of a total of 324 used subbands. The spectral behaviour of the amplitude and phase solutions can give us insight into whether we are absorbing true flux or calibration artefacts into our model: if our calibration model becomes more accurate to the true underlying flux, we expect e.g. the phase of core station calibration solutions to remain the same (as we are improving high-resolution details of our model) while our international station calibration solutions would become flatter in amplitude. Similarly, we expect to see large-scale frequency-dependent behaviour in the phase solutions of international stations, with smaller fluctuations around this behaviour: the effect of differential TEC, ionospheric turbulence, etc. Indeed, while the core stations (and, to a lesser extent, the remote stations) can be approximated as having the ``same ionosphere" above them, this is not the case for international stations: this is a major source of the difficulties in calibrating these stations. As such, for international stations, we expect to see complex structure in the phase gain solutions, characteristic of the differences between the local atmosphere above the core and the international stations. For core stations, we only expect to see structure characteristic of scintillation (regular stripes with no frequency-dependent structure) in the phase gain solutions.

\pg
\cref{fig.fullBW.CS005,fig.fullBW.CS201HBA0} show the gain solution phases and amplitudes for all times and frequencies in our observation. Some patches of the plot are empty (horizontal red bands): this is because some subbands had very poor data, and their use in self-calibration corrupted the final image. They are more numerous for the first calibration pass than the last, as the solutions were not sought for some cases.

@@

\pg
We see that some of the structure in the amplitude gains shown in \cref{fig.fullBW.DE601HBA} goes away over the course of self-calibration, though one of the two arcs present in the plot appears to decrease in amplitude. For \cref{fig.fullBW.DE602HBA}, we see that the pattern in the amplitude plot clearly changes - the fact that the pattern in \cref{fig.fullBW.DE602HBA.before} does not appear to be either dispersion or scintillation pattern would indicate that it consists of unmodeled flux. While some structure remains in the last 2 hours and low frequencies for this antenna's gain solution amplitudes, the last 2 hours of the observation are generally poor compared to the first six: we expect that the quality-based weighting scheme will minimise the impact of these poor calibration solutions on the final images. In \cref{fig.fullBW.DE603HBA}, we see that the self-calibration process was clearly less effective than it was for DE602HBA: while some of the structure was minimised (in the first two hours particularly: we go from the peculiar ``ellipse" pattern to something which looks more like scintillation), a lot clearly remains. Our self-calibration process could therefore likely be taken further. However, in our last step, we use 324 out of the full 366 subbands (leaving some out due to poor quality); a single round of self-calibration with this full dataset takes 5000 minutes on the best computing cluster available for this work. Because a single round of self-calibration would, at this point, take as much time as imaging a portion of the EGS due to computational constraints, we treat the self-calibration process as ``having converged" at this point: while it could be yet improved, we have reached a signal-to-noise of 2.78mJy.

@@

\pg
The final, high-resolution, multi-frequency, bootstrapped model of 3C295 - which gives a spectral index for each pixel in the model - is shown in \cref{fig.3c295.final}. An overlay of this model is shown in \cref{fig.3c295.final.overlay}. The brightness distribution seems to have largely stayed within the bounds of the initial model for the southern lobe, but to have shifted eastwards for the northern lobe. We see that artefacts remain in the field, but this is not the case at all frequencies, as shown in \cref{fig.3c295.spectral.overlays}: three of the six slices seem to be much better-calibrated than their peers. It should be noted that we both maximise the presence of calibration artefacts \textit{and} minimise the contribution of diffuse emission through our choices of weighting scheme and imaging $uv$-cut, respectively.

@@

\begin{figure}[h]
\includegraphics[width=0.9\linewidth]{images/{3C_295_I_8.7GHz_lbs2003.J2000.fits.overlay}.png}
\caption{\label{fig.3c295.final.overlay} Overlay of final model of 3C295 on initial VLA model. This final model is made by adding individual contributions from the spectral slices shown in \cref{fig.3c295.spectral.overlays}.}
\end{figure}
% python ~/Scripts/PngCutoutFromFits.py --filename 3C_295_I_8.7GHz_lbs2003.J2000.fits --RA 14:11:20.6 --Dec 52:12:09.0 --size .1 --invert -v --overlaymin 10 --nlevels 200 --vmin 0.06 --vmax 2 --overlay IMAGES/MSlist.fullBW.LOBOS6.MSlist.fullBW.txt.uniform.int.restored.bootstrap.fits
\pg
By inspecting individual spectral slices of our multi-spectral model in \cref{fig.3c295.spectral.overlays}, we can see that the presence of calibration artefacts is not the same at all frequencies. We see a lot of artefacts remain for the slice at 176.17 MHz in particular, and the slice at 140.428 MHz to a lesser extent. Note that the overlay values rise exponentially. The maximum pixel value  in the image is $\sim$20 Jy, while the noise in the corner of the image is $\sim$2mJy. We achieve a relatively decent dynamic range (DR=10000, which is a factor of 100 lower than what is achievable with self-cal on VLA, WSRT data). We note that the dynamic range here is lower than could be expected for this data. This could be due, among other things, to our choices of $uv$-cut and imaging weights, both of which ``artificially" increase the rms in the image (but are necessary to improve image deconvolution for such a diffuse source. However, considering the still-large presence of artefacts near 3C295, it is likely that we ended up in a local minimum by not modeling all the sources in the field.

@@

\pg
It is interesting to note that the rms value in each individual slice is comparable to the rms value in the ``collapsed" model (average of all slices), where we would expect it would decrease by a factor $\sim\sqrt{6}$ (i.e. about 2.5). The fact that this does not occur is a strong indication that we are in a regime where we are dominated by calibration artefacts, and that the rms is therefore not truly thermal. It is also an indication that we could likely improve the self-calibration strategy: this behaviour could be due to too small a deconvolution mask, for example.

@@

\pg
Because of time/computational constraints, we used calibration solution intervals of 8 channels and 1 minute in time and frequency respectively. Given the brightness of 3C295, this was clearly not justified on the basis of signal-to-noise; it did, however, mean that a single pass of self-calibration on one subband took 10 minutes rather than an hour and a half (factor of 9 improvement in time). With the advent of new tools like cubical \citepads{2018MNRAS.tmp.1165K,2018ascl.soft05031K}, a much faster direction-independent implementation of the algorithm used for killMS\footnote{In the context of this work, it would literally be an equivalent but faster implementation of exactly the same algorithm.} which was in development throughout my PhD, it is likely possible to dramatically speed the calibration part of the self-calibration loop (the imaging time constraint would dominate as more subbands are added). Because of the use of quality-based weighting schemes, it is unlikely that a dramatic improvement would be seen when self-calibrating over a few subbands, as the relative contribution of poorly-calibrated visibilities would be down-weighted by the quality-based weighting scheme: as more and more data is added, however, the relative contribution of the artefacts introduced by the choice of calibration intervals would likely begin to overcome the purely thermal noise in the images, which is expected to decrease as more data is added.% in the well-calibrated visibilities, even when down-weighted.

\pg
All the same, this project served as a useful exercise in both developing a good working methodology and in testing calibration strategies for international LOFAR datasets. Much of the development which went into this project and led to no directly useful results still proved an extremely helpful context in which to understand how to improve the LOFAR pipeline. In particular, gain smoothing has shown promising result on LOTSS images, but was not used in this project because of the full-Jones solutions: properly implementing this smoothing in polarisation could provide the kind of regularisation that our gain solutions need.

>>>>

\pg
The resulting model is expected to be reliable enough to calibrate our Groth Strip data for the full LOFAR array and the entire HBA bandwidth. It would theoretically be possible to use the direction-dependent gain solutions from \cref{section.EGS.lowres} to improve our images, but these are only available for core and remote stations, and not international stations - direction-dependent effects are therefore expected to still arise in the final images due to this. As such, we focus on direction-independent calibration from this point on. Quantifying the impact of the resulting directional gain errors is the subject of \cref{section.decorr}.

@@

\pg
As outlined in \cref{section.RIME}, calibration is the process of solving for Jones matrices, which are associated to individual antennas. Here, the only source in our model (the source that dominates, by far, over other sources in the field) is 3C295; we can thus limit ourselves to \textit{direction-independent} calibration, which is equivalent to performing direction-dependent calibration in the direction of 3C295 only. The associated RIME can be written as:
\begin{align}
\Vis_{pq}^{\nu} &= \Gjones_p^{\nu} \left( \sum_{s=3C295} \Ejones_{sp}^{\nu} \Kjones_{sp}^{\nu} \Bmatrix \left(\Kjones_{sq}^{\nu}\right)^H \left(\Ejones_{sq}^{\nu}\right)^H \right) \left(\Gjones_q^{\nu}\right)^H\\
		  &= \Gjones_p^{\nu} \Ejones_{3C295,p}^{\nu} \Bmatrix_{3C295}^{\nu} k_{s,pq}\left(\Ejones_{3C295,q}^{\nu} \Gjones_q^{\nu}\right)^H\\
		  &= \Jones_{3C295,p}^{\nu} \Bmatrix_{3C295}^{\nu}k_{s,pq} \left(\Jones_{3C295,q}^{\nu}\right)^H\label{3c295.dirin.cal}
\end{align}
where we mute the time-dependence of all terms, and explicitly model the LOFAR beam during both calibration and imaging, so as to ensure that the true sensitivity of each baseline to this out-of-phase-centre calibrator is taken into account appropriately.

\pg
What interests us here is therefore the $\Jones_{3C295,p}^{\nu}$ for all international stations $p$ and at all times and frequencies $t,\nu$. We improve conditioning by using time and frequency solution intervals of a minute and 8 subbands; at those times where this interval is too large, the quality-based weighting scheme is expected to compensate. Choosing such large intervals also helps self-calibration converge faster. We do not explicitly resort to physics-based solutions to constrain the calibration solutions: we solve for the full 8 parameters of each complex $2\times 2$ Jones matrix (2 parameters per complex quantity, 4 complex quantities).

@@

\pg
Because we apply a high-pass $uv$-cut of 1km when calibrating, and a $uv$-cut of 100km when imaging\footnote{i.e. all baselines shorter than 100km are discarded.}, we need to make sure that we do not introduce spurious structure in the core and remote station gains. Similarly, we can check the quality of our self-calibration by looking at the structure of the international station gains. Gain structure can consist of genuine atmospheric \& instrumental effects, but can also include unmodeled flux being absorbed into the gain solutions. At first, this is unavoidable: if flux is resolved by a baseline but not included in the model, then it will be absorbed into the gains. As we improve our model, we expect the gain curves to flatten and become smoother. If this happens for our international station gains, then we have a strong indication that we are in fact recovering physical signal rather than adding artefacts into our model. Similarly, we expect core and remote station gains to remain mostly static: if this is the case, then our strategy does not bias our gain solutions for these stations away from the underlying truth.

@@

\pg
In so doing, we introduce a frequency-dependent error factor in our gain solutions due to the fact that the absolute flux scale isn't known \textit{a priori}:
\begin{align}
\Vis_{pq}^\nu&=\Jones_{s,p}^\nu \Bmatrix \left(\Jones_{s,q}^\nu\right)^H\\
         &=\Jones_{s,p}\sqrt{\alpha_\nu} \frac{\Bmatrix}{\alpha_\nu} \left(\Jones_{s,q}^\nu\sqrt{\alpha_\nu}\right)^H\\
         &=\hatJones_{3C295,p}^{\nu}\frac{\Bmatrix}{\alpha_\nu} \left(\hatJones_{s,q}^\nu\right)^H\\
\hatJones_{3C295,p}^{\nu} &= \sqrt{\alpha_\nu} \Jones_{3C295,p}^{\nu}
\end{align}
where $\hatJones_{3C295,p}^{\nu}$ is our estimation of the Jones matrix $\Jones_{3C295,p}^{\nu}$ for time $t$ and frequency $\nu$, and $\alpha_\nu$ is a smoothly-varying  scalar function of frequency alone. This then introduces a frequency-dependent error factor $1/\sqrt{\alpha_\nu}$ in our visibilities and therefore in our images. This factor is a constant for all visibilities at a given frequency.

@@

% images:
% IMAGES/3c295.fromVLA.6sb.withbeam.pass[1-5].int.model.fits IMAGES/3c295.fromVLA.60sb.withbeam.pass[1-3].int.model.fits IMAGES/MSlist.fullBW.LOBOS6.MSlist.fullBW.txt.uniform.int.model.fits
% script: /home/offler/plotSCamp.py
\begin{figure}[h!]
\includegraphics[width=0.8\linewidth]{images/AmpFluct.png}
\caption{\label{fig.SC.ampfluct} Amplitude fluctuation of the models resulting from self-calibration iterations, expressed as a ratio of the model's integrated flux divided by the Scaife \& Healde integrated flux model at the appropriate frequency. Note that the integrated model flux for 324sb-pass1 (the penultimate data point in this plot) was lost due to a careless command: the value shown here is just the average of the two neighbouring values - i.e. those of 60sb-pass3 and 324sb-pass2.}
\end{figure}

\pg
Because $\alpha$ is smooth in frequency, we can apply a frequency-dependent amplitude correction (i.e. ``bootstrap") by relying on the work of Scaife \& Heald \citep[see][]{arse}. This correction will be done in the imaging plane, since $alpha$ is a function of $uv$-position (but not sky position). This is, again, because 3C295 does not lie at phase centre: if it did, the correction could be done equivalently in visibility or image space. By performing it in the image-plane, we rid ourselves of the $uv$-dependence. The resulting signal-to-noise improvement is shown in \cref{fig.SCrms}. It is calculated on the bootstrapped residual images resulting from each round of the self-calibration process - as such, the only thing that changes is the noise, since the signal is corrected to be the appropriate constant As we can see, the noise falls nicely with self-calibration passes. Tragically, the results from the first full-bandwidth calibration run with the 60-subband model on the full 324 good subbands were lost before this plot could be made, and so only the results from the second full-bandwidth self-calibration pass remain. We can see that it is a small improvement over the 60-subband model. 
\begin{figure}[h!]
\includegraphics[width=0.8\linewidth]{images/SCrms.png}
\caption{\label{fig.SCrms} Bootstrapped residual RMS for each calibration pass. Note that the data for 324sb-pass1 (the penultimate data in this plot) were lost due to a careless command: the value shown here is just the average of the two neighbouring values.}
\end{figure}

\pg
The large decrease (factor of $\sim2$) in RMS between the last 6-subband residual image and the first 60-subband residual image in \cref{fig.SCrms} is significant: we would expect a similar jump when going from 60 to 324 subbands. The fact that we do not is a strong indicator that we are digging into the limits of what our strategy can achieve by the end of the second full-bandwidth self-calibration pass: we are stuck in a local minimum. This is likely due to the fact that our initial sky model (and all subsequent models) only include 3C295: the fact that all the other sources in the primary beam are not included in the model is bound to introduce bias our gain solutions and artefacts in our images.

@@

\subsubsection{Gains for 6 antennas at a given frequency}

\pg
Let us begin by checking whether our core and remote station gains change over time due to our self-calibration procedure. Here, we look at the gains of SB155, on which we solved for gain solutions using our initial and final models. This subband was included in both the 60-subband self-calibration step and the 6-subband self-calibration step.
\begin{figure}[h!]
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.CS005HBA0.firstpass}.png}
\caption{\label{fig.sb155.gains.cs201.before} CS005, first calibration}
\end{subfigure}
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.CS005HBA0.lastpass}.png}
\caption{\label{fig.sb155.gains.cs201.after} CS005, last calibration}
\end{subfigure}
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.RS305HBA.firstpass}.png}
\caption{\label{fig.sb155.gains.RS305.before} RS305, first calibration}
\end{subfigure}
\begin{subfigure}{.9\textwidth}
\includegraphics[width=\linewidth]{images/{SB155.RS305HBA.lastpass}.png}
\caption{\label{fig.sb155.gains.RS305.afer} RS305, last calibration}
\end{subfigure}
\caption{\label{fig.sb155.gains.CR} Gain amplitude \& phase for CS201 of SB155}
\end{figure}

@@

\subsubsection{Spectral behaviour of gain solutions}

\pg
So far, we have shown the gain solutions for one subband, out of a total of 324 used subbands. The spectral behaviour of the amplitude and phase solutions can give us insight into whether we are absorbing true flux or calibration artefacts into our model: if our calibration model becomes more accurate to the true underlying flux, we expect e.g. the phase of core station calibration solutions to remain the same (as we are improving high-resolution details of our model) while our international station calibration solutions would become flatter in amplitude. Similarly, we expect to see large-scale frequency-dependent behaviour in the phase solutions of international stations, with smaller fluctuations around this behaviour: the effect of differential TEC, ionospheric turbulence, etc. Indeed, while the core stations (and, to a lesser extent, the remote stations) can be approximated as having the ``same ionosphere" above them, this is not the case for international stations: this is a major source of the difficulties in calibrating these stations. As such, for international stations, we expect to see complex structure in the phase gain solutions, characteristic of the differences between the local ionosphere above the core and the international stations. For core stations, we only expect to see structure characteristic of scintillation (regular stripes with no frequency-dependent structure) in the phase gain solutions.

\pg
\cref{fig.fullBW.CS005,fig.fullBW.CS201HBA0} show the gain solution phases and amplitudes for all times and frequencies in our observation. Some patches of the plot are empty (horizontal black bands): this is because some subbands had very poor data, and their use in self-calibration corrupted the final image. They are more numerous for the first calibration pass than the last, as the solutions were not sought for some cases.

@@

\pg
We see that some of the structure in the amplitude gains shown in \cref{fig.fullBW.DE601HBA} goes away over the course of self-calibration, though one of the two arcs present in the plot appears to decrease in amplitude. For \cref{fig.fullBW.DE602HBA}, we see that the pattern in the amplitude plot clearly changes - the fact that the pattern in \cref{fig.fullBW.DE602HBA.before} does not appear to be either dispersion or scintillation pattern would indicate that it consists of unmodeled flux. While some structure remains in the last 2 hours and low frequencies for this antenna's gain solution amplitudes, the last 2 hours of the observation are generally poor compared to the first six: we expect that the quality-based weighting scheme will minimise the impact of these poor calibration solutions on the final images. In \cref{fig.fullBW.DE603HBA}, we see that the self-calibration process was clearly less effective than it was for DE602HBA: while some of the structure was minimised (in the first two hours particularly: we go from the peculiar ``ellipse" amplitude pattern to something which looks more like scintillation), a lot clearly remains. Our self-calibration process could therefore likely be taken further. However, in our last step, we use 324 out of the full 366 subbands (leaving some out due to poor quality); a single round of self-calibration with this full dataset takes 5000 minutes on the best computing cluster available for this work. Because a single round of self-calibration would, at this point, take as much time as imaging a portion of the EGS due to computational constraints, we treat the self-calibration process as ``having converged" at this point: while it could be yet improved, we have reached a signal-to-noise ratio of 2.78 mJy.

@@

\pg
The final, high-resolution, multi-frequency, bootstrapped model of 3C295 - which gives a spectral index for each pixel in the model - is shown in \cref{fig.3c295.final}. An overlay of this model is shown in \cref{fig.3c295.final.overlay}. The brightness distribution seems to have largely stayed within the bounds of the initial model for the southern lobe, but to have shifted eastwards for the northern lobe. We see that artefacts remain in the field, but this is not the case at all frequencies, as shown in \cref{fig.3c295.spectral.overlays}: three of the six slices (a,b,d) seem to be much better-calibrated than their peers. It should be noted that we both maximise the presence of calibration artefacts \textit{and} minimise the contribution of diffuse emission through our choices of weighting scheme and imaging $uv$-cut, respectively.

@@

\begin{figure}[h]
\includegraphics[width=0.9\linewidth]{images/{3C_295_I_8.7GHz_lbs2003.J2000.fits.overlay}.png}
\caption{\label{fig.3c295.final.overlay} Overlay of final model of 3C295 on initial VLA model. This final model is made by adding individual contributions from the spectral slices shown in \cref{fig.3c295.spectral.overlays}.}
\end{figure}
% python ~/Scripts/PngCutoutFromFits.py --filename 3C_295_I_8.7GHz_lbs2003.J2000.fits --RA 14:11:20.6 --Dec 52:12:09.0 --size .1 --invert -v --overlaymin 10 --nlevels 200 --vmin 0.06 --vmax 2 --overlay IMAGES/MSlist.fullBW.LOBOS6.MSlist.fullBW.txt.uniform.int.restored.bootstrap.fits
\pg
By inspecting individual spectral slices of our multi-spectral model in \cref{fig.3c295.spectral.overlays}, we can see that the presence of calibration artefacts is not the same at all frequencies. We see a lot of artefacts remain for the slice at 176.17 MHz in particular, and the slice at 140.428 MHz to a lesser extent. Note that the overlay values rise exponentially. The maximum pixel value  in the image is $\sim$20 Jy, while the noise in the corner of the image is $\sim$2 mJy. We achieve a relatively decent dynamic range (DR=10000, which is a factor of 100 lower than what is achievable with self-cal on VLA, WSRT data). We note that the dynamic range here is lower than could be expected for this data. This could be due, among other things, to our choices of $uv$-cut and imaging weights, both of which ``artificially" increase the rms in the image but are necessary to improve image deconvolution for such a diffuse source. However, considering the still-large presence of artefacts near 3C295, it is likely that we ended up in a local minimum by not modeling all the sources in the field.

@@

\pg
It is interesting to note that the rms value in each individual slice is comparable to the rms value in the ``collapsed" model (average of all slices), where we would expect a decrease by a factor $\sim\sqrt{6}$ (i.e. about 2.5). The fact that this does not occur is a strong indication that we are in a regime where we are dominated by calibration artefacts, and that the rms is therefore not truly thermal. It is also an indication that we could likely improve the self-calibration strategy: this behaviour could be due to too small a deconvolution mask, for example.

@@

\pg
Because of time/computational constraints, we used calibration solution intervals of 8 channels and 1 minute in time and frequency respectively. Given the brightness of 3C295, this was clearly not justified on the basis of signal-to-noise; it did, however, mean that a single pass of self-calibration on one subband took 10 minutes rather than the hour and a half with solution intervals of 4 channels and 8 seconds in time and frequency respectively (factor of 9 improvement in time). With the advent of new tools like cubical \citepads{2018MNRAS.tmp.1165K,2018ascl.soft05031K}, a much faster direction-independent implementation of the algorithm used for killMS\footnote{In the context of this work, it would literally be an equivalent but faster implementation of exactly the same algorithm.} which was in development throughout my PhD, it is likely possible to dramatically speed the calibration part of the self-calibration loop (the imaging time constraint would dominate as more subbands are added). Because of the use of quality-based weighting schemes, it is unlikely that a dramatic improvement would be seen when self-calibrating over a few subbands, as the relative contribution of poorly-calibrated visibilities would be down-weighted by the quality-based weighting scheme: as more and more data is added, however, the relative contribution of the artefacts introduced by the choice of calibration intervals would likely begin to overcome the purely thermal noise in the images, which is expected to decrease as more data is added.% in the well-calibrated visibilities, even when down-weighted.

\pg
All the same, this project served as a useful exercise in both developing a good working methodology and in testing calibration strategies for international LOFAR datasets. Much of the development which went into this project and led to no directly useful results still proved an extremely helpful context in which to understand how to improve the LOFAR pipeline. In particular, gain smoothing\footnote{i.e. the application of a smoothing operation in gain phase and amplitude, justified by the fact that they are expected to be smooth and so jagged behaviour is likely erroneous.} has shown promising result on LOTSS images, but was not used in this project because of the full-Jones solutions: properly implementing this smoothing in a framework capable of accounting for polarisation could provide the kind of regularisation that our gain solutions need.
